\textbf{Effective application of machine learning to a microbiome-based classification problem}

### 2020-01-17 submission/manuscript.tex
some additions for rebuttal

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Effective application of machine learning to microbiome-based classification problems},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{\textbf{Effective application of machine learning to microbiome-based
classification problems}}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

@@

@@

Among microbiome researchers, the lack of justification when selecting a
modeling approach has often been due to an implicit assumption that more
complex models are better. This has resulted in a trend towards using
non-linear models such as random forest and deep neural networks (3, 12,
25--27) over simpler models such as logistic regression or other linear
models (19, 23, 28). Although in some cases, complex models may capture
important non-linear relationships and therefore yield better
predictions, they can also result in black boxes that lack
interpretability. Such models require post hoc explanations to quantify
the importance of each feature in making predictions. Depending on the
goal of the modeling, other approaches may be more appropriate. For
example, researchers trying to identify the microbiota associated with
disease may desire a more interpretable model, whereas clinicians may
emphasize predictive performance. Nonetheless, it is essential to
understand that the benefit of more complex, less interpretable models
may be minimal (29, 30). It is important for researchers to justify
their choice of modeling approach.

To showcase a rigorous ML pipeline and to shed light on how ML model
selection can affect modeling results, we performed an empirical
analysis comparing seven modeling approaches with the same dataset and
pipeline. We built three linear models with different forms of
regularization: L2-regularized logistic regression and L1 and
L2-regularized support vector machines (SVM) with a linear kernel. We
also trained four non-linear models: SVM with radial basis function
kernel, a decision tree, random forest, and gradient boosted trees. We
compared their predictive performance, interpretability, and training
time. To demonstrate the performance of these modeling approaches and
our pipeline, we used data from a previously published study that sought
to classify individuals as having healthy colons or colonic lesions
based on the 16S rRNA gene sequences collected from fecal samples (4).
This dataset was selected because it is a relatively large collection of
individuals (N=490) connected to a clinically significant disease where
there is ample evidence that the disease is driven by variation in the
microbiome (2, 4, 5, 31). With this dataset, we developed an ML pipeline
that can be used in many different scenarios for training and evaluating
models. This framework can be easily applied to other host-associated
and environmental microbiome datasets.

@@

We trained seven different models using the training data {[}Table 1{]}.
We focused on different classification algorithms and regularization
methods. Regularization helps to prevent overfitting by penalizing a
model that fits the training data too well (32). For regularized
logistic regression and SVM with a linear kernel, we used
L2-regularization to keep all potentially important features. For
comparison, we also trained an L1-regularized SVM with a linear kernel.
L1-regularization on microbiome data led to a sparser solution (i.e.,
forced many coefficients to zero). To explore the potential for
non-linear relationships among features to improve classification, we
trained tree-based models including, a decision tree, a random forest,
and gradient boosted trees (XGBoost) and an SVM with a non-linear
kernel.

@@

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performance of the seven models to
classify individuals as having healthy colons or SRNs {[}Figure 2{]}.
The predictive performance of random forest model was higher than other
ML models with a median 0.695 {[}IQR 0.650-0.739{]}, though not
significantly (p=0.5) (Figure S3). Similarly, L2-regularized logistic
regression, XGBoost, L2-regularized SVM with linear and radial basis
function kernel AUROC values were not significantly different from one
another and had median AUROC values of 0.680 {[}IQR 0.639-0.750{]},
0.679 {[}IQR 0.643-0.746{]}, 0.678 {[}IQR 0.639-0.750{]} and 0.668
{[}IQR 0.639-0.750{]}, respectively. L1-regularized SVM with linear
kernel and decision tree had significantly lower AUROC values than the
other ML models with median AUROC of 0.650 {[}IQR 0.629-0.760{]} and
0.601 {[}IQR 0.636-0.753{]}, respectively {[}Figure 2{]}. Interestingly,
these results demonstrate that the most complex model (XGBoost) did not
have the best performance and that the most interpretable models
(L2-regularized logistic regression and L2-regularized SVM with linear
kernel) performed nearly as well as non-linear models.

@@

\textbf{Interpretation of each ML model.} Interpretability is related to
the degree to which humans can understand the reasons behind a model
prediction (33--35). Because we often use ML models not just to predict
a health outcome, but also to identify potential biomarkers for disease,
model interpretation becomes crucial for microbiome studies. ML models
often decrease in interpretability as they increase in complexity. In
this study, we used two methods to help interpret our models.

@@

Second, to analyze non-linear models, we interpreted the feature
importance using permutation importance (36). Whereas the absolute
feature weights were determined from the trained models, here we
measured importance using the held-out test data. Permutation importance
analysis is a post hoc explanation of the model, in which we randomly
permuted groups of perfectly correlated features together and other
features individually across the two groups in the held-out test data.
We then calculated how much the predictive performance of the model
(i.e., testing AUROC values) decreased when each OTU or group of OTUs
was randomly permuted. We ranked the OTUs based on how much the median
testing AUROC decreased when it was permuted; the OTU with the largest
decrease ranked highest {[}Figure 4{]}. Among the twenty OTUs with the
largest impact, there was only one OTU (OTU 822) that was shared among
all of the models; however, we found three OTUs (OTUs 58, 110, 367) that
were important in each of the tree-based models. Similarly, the random
forest and XGBoost models shared four of the most important OTUs (OTUs
2, 12, 361, 477). Permutation analysis results also revealed that with
the exception of the decision tree model, removal of any individual OTU
had minimal impact on model performance. For example, if OTU 367 was
permuted across the samples in the decision tree model, the median AUROC
dropped from 0.601 to 0.525. In contrast, if the same OTU was permuted
in the random forest model, the AUROC only dropped from 0.695 to 0.680,
which indicated high degree of collinearity in the dataset. Permutation
analysis allowed us to gauge the importance of each OTU in non-linear
models and partially account for collinearity by grouping correlated
OTUs to determine their impact as a group.

@@

We highlighted the importance of model interpretation to gain greater
biological insights into microbiota-associated diseases. In this study,
we showcased two different interpretation methods: ranking each OTU by
(i) their absolute weights in the trained models and (ii) their impact
on the predictive performance based on permutation importance.
Human-associated microbial communities have complex correlation
structures that create collinearity in the datasets. This can hinder our
ability to reliably interpret models because the feature weights of
correlated OTUs are influenced by one another (37). To capture all
important features, once we identify highly ranked OTUs, we should
review their relationships with other OTUs. These relationships will
help us generate new hypotheses about the ecology of the disease and
test them with follow-up experiments. When we used permutation
importance, we partially accounted for collinearity by grouping
correlated OTUs to determine their impact as a group. We grouped OTUs
that had a perfect correlation with each other; however, we could reduce
the correlation threshold to further investigate the relationships among
correlated features. It is important to know the correlation structures
of the data to avoid misinterpreting the models. This is likely to be a
particular problem with shotgun metagenomic datasets where collinearity
will be more pronounced due to many genes being correlated with one
another because they come from the same chromosome. To identify the true
underlying microbial factors of a disease, it is crucial to follow up on
any correlation analyses with further hypothesis testing and
experimentation for biological validation.

In this study, we did not consider all possible modeling approaches.
However, the principles highlighted throughout this study apply to other
ML modeling tasks with microbiome data. For example, we did not evaluate
multicategory classification methods to predict non-binary outcomes. We
could have trained models to differentiate between people with healthy
colons and those with adenomas or carcinomas (k=3 categories). We did
not perform this analysis because the clinically relevant diagnosis
grouping was between patients with healthy colons and those with SRNs.
Furthermore, as the number of classes increases, more samples are
required for each category to train an accurate model. We also did not
use regression-based analyses to predict a non-categorical outcome. We
have previously used such an approach to train random forest models to
predict fecal short-chain fatty acid concentrations based on microbiome
data (38). Our analysis was also limited to shallow learning methods and
did not explore deep learning methods such as neural networks. Deep
learning methods hold promise (12, 39, 40), but microbiome datasets
often suffer from having many features and small sample sizes, which
result in overfitting.

Our framework provides a reproducible structure to investigators wanting
to train, evaluate, and interpret their own ML models to generate
hypotheses regarding which OTUs might be biologically relevant. However,
deploying microbiome-based models to make clinical diagnoses or
predictions is a significantly more challenging and distinct undertaking
(41). For example, we currently lack standardized methods to collect
patient samples, generate sequence data, and report clinical data. We
are also challenged by the practical constraints of OTU-based
approaches. The de novo algorithms commonly in use are slow, require
considerable memory, and result in different OTU assignments as new data
are added (42). Finally, we also need independent validation cohorts to
test the performance of a diagnostic model. To realize the potential for
using ML approaches with microbiome data, it is necessary that we direct
our efforts to overcome these challenges.

@@

\textbf{Data collection and study population.} The original stool
samples described in our analysis were obtained from patients recruited
by Great Lakes-New England Early Detection Research Network (5). Stool
samples were provided by adults who were undergoing a scheduled
screening or surveillance colonoscopy. Participants were recruited from
Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor
(MI, USA). Patients' colonic health was visually assessed by colonoscopy
with bowel preparation and tissue histopathology of all resected
lesions. We assigned patients into two classes: those with healthy
colons and those with screen relevant neoplasias (SRNs). The healthy
class included patients with healthy colons or non-advanced adenomas,
whereas the SRN class included patients with advanced adenomas or
carcinomas (43). Patients with an adenoma greater than 1 cm, more than
three adenomas of any size, or an adenoma with villous histology were
classified as having advanced adenomas (43). There were 172 patients
with normal colonoscopies, 198 with adenomas, and 120 with carcinomas.
Of the 198 adenomas, 109 were identified as advanced adenomas. Together
261 patients were classified as healthy and 229 patients were classified
as having an SRN.

\textbf{16S rRNA gene sequencing data.} Stool samples provided by the
patients were used for 16S rRNA gene sequencing to measure bacterial
population abundances. The sequence data used in our analyses were
originally generated by Baxter et al. (available through NCBI Sequence
Read Archive {[}SRP062005{]}, (5)). The OTU abundance table was
generated by Sze et al (44), who processed the 16S rRNA sequences in
mothur (v1.39.3) using the default quality filtering methods,
identifying and removing chimeric sequences using VSEARCH, and assigning
to OTUs at 97\% similarity using the OptiClust algorithm (42, 45, 46);
(\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
These OTU abundances were the features we used to predict colorectal
health of the patients. There were 6920 OTUs. OTU abundances were
subsampled to the size of the smallest sample and normalized across
samples such that the highest abundance of each OTU would be 1, and the
lowest would be 0.

@@

For hyperparameter selection, we started with a granular grid search.
Then we narrowed and fine-tuned the range of each hyperparameter.For
L2-regularized logistic regression, L1 and L2-regularized SVM with
linear and radial basis function kernels, we tuned the cost
hyperparameter, which controls the regularization strength, where
smaller values specify stronger regularization. For SVM with radial
basis function kernel, we also tuned the sigma hyperparameter, which
determines the reach of a single training instance where, for a high
value of sigma, the SVM decision boundary will be dependent on the
points that are closest to the decision boundary. For the decision tree
model, we tuned the depth of the tree where the deeper the tree, the
more splits it has. For random forest, we tuned the number of features
to consider when looking for the best tree split. For XGBoost, we tuned
the learning rate and the fraction of samples used for fitting the
individual base learners. Performing a grid search for hyperparameter
selection might not be feasible for when there are more than two
hyperparameters to tune for. In such cases, it is more efficient to use
random search or recently developed tools such as Hyperband to identify
good hyperparameter configurations (47).

@@

\hypertarget{ref-knights_human-associated_2011}{}
31. \textbf{Knights D}, \textbf{Parfrey LW}, \textbf{Zaneveld J},
\textbf{Lozupone C}, \textbf{Knight R}. 2011. Human-associated microbial
signatures: Examining their predictive value. Cell Host Microbe
\textbf{10}:292--296.
doi:\href{https://doi.org/10.1016/j.chom.2011.09.003}{10.1016/j.chom.2011.09.003}.

\hypertarget{ref-Ng:2004:FSL:1015330.1015435}{}
32. \textbf{Ng AY}. 2004. Feature selection, l1 vs. l2 regularization,
and rotational invariance, pp. 78. \emph{In} Proceedings of the
twenty-first international conference on machine learning. ACM, New
York, NY, USA.

\hypertarget{ref-miller_explanation_2017}{}
33. \textbf{Miller T}. 2017. Explanation in artificial intelligence:
Insights from the social sciences. arXiv:170607269 {[}cs{]}.

\hypertarget{ref-ribeiro_why_2016}{}
34. \textbf{Ribeiro MT}, \textbf{Singh S}, \textbf{Guestrin C}. 2016.
``Why should i trust you?'': Explaining the predictions of any
classifier. arXiv:160204938 {[}cs, stat{]}.

\hypertarget{ref-nori_interpretml:_2019}{}
35. \textbf{Nori H}, \textbf{Jenkins S}, \textbf{Koch P},
\textbf{Caruana R}. 2019. InterpretML: A unified framework for machine
learning interpretability. arXiv:190909223 {[}cs, stat{]}.

\hypertarget{ref-10.1093ux2fbioinformaticsux2fbtq134}{}
36. \textbf{Altmann A}, \textbf{Toloşi L}, \textbf{Sander O},
\textbf{Lengauer T}. 2010. Permutation importance: a corrected feature
importance measure. Bioinformatics \textbf{26}:1340--1347.
doi:\href{https://doi.org/10.1093/bioinformatics/btq134}{10.1093/bioinformatics/btq134}.

\hypertarget{ref-dormann_collinearity:_2013}{}
37. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
\textbf{Buchmann C}, \textbf{Carl G}, \textbf{Carré G}, \textbf{Marquéz
JRG}, \textbf{Gruber B}, \textbf{Lafourcade B}, \textbf{Leitão PJ},
\textbf{Münkemüller T}, \textbf{McClean C}, \textbf{Osborne PE},
\textbf{Reineking B}, \textbf{Schröder B}, \textbf{Skidmore AK},
\textbf{Zurell D}, \textbf{Lautenbach S}. 2013. Collinearity: A review
of methods to deal with it and a simulation study evaluating their
performance. Ecography \textbf{36}:27--46.
doi:\href{https://doi.org/10.1111/j.1600-0587.2012.07348.x}{10.1111/j.1600-0587.2012.07348.x}.

\hypertarget{ref-sze_fecal_2019}{}
38. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
\textbf{Ruffin MT}, \textbf{Schloss PD}. 2019. Fecal short-chain fatty
acids are not predictive of colonic tumor status and cannot be predicted
based on bacterial community structure. mBio \textbf{10}:e01454--19.
doi:\href{https://doi.org/10.1128/mBio.01454-19}{10.1128/mBio.01454-19}.

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\hypersetup{
            pdftitle={Effective application of machine learning to a microbiome-based classification problem},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

@@

\title{\textbf{Effective application of machine learning to a microbiome-based
classification problem}}
\author{}
\date{\vspace{-2.5em}}

@@

Among microbiome researchers, the lack of justification when selecting a
modeling approach has often been due to an implicit assumption that more
complex models are better. This has resulted in a trend towards using
non-linear models such as random forest and deep neural networks (3, 12,
25--27) over simpler models such as logistic regression or other linear
models (19, 23, 28). Although in some cases, complex models may capture
important non-linear relationships and therefore yield better
predictions, they can also result in black boxes that lack
interpretability. Such models require post hoc explanations to quantify
the importance of each feature in making predictions. Depending on the
goal of the modeling, other approaches may be more appropriate. For
example, researchers trying to identify the microbiota associated with
disease may desire a more interpretable model, whereas clinicians may
emphasize predictive performance. Nonetheless, it is essential to
understand that the benefit of more complex, less interpretable models
may be minimal (29--31). It is important for researchers to justify
their choice of modeling approach.

In this study, we provided steps toward standardization of machine
learning methods for microbiome studies which are often poorly
documented and executed. To showcase a rigorous ML pipeline and to shed
light on how ML model selection can affect modeling results, we
performed an empirical analysis comparing seven modeling approaches with
the same dataset and pipeline. We built three linear models with
different forms of regularization: L2-regularized logistic regression
and L1 and L2-regularized support vector machines (SVM) with a linear
kernel. We also trained four non-linear models: SVM with radial basis
function kernel, a decision tree, random forest, and gradient boosted
trees. We compared their predictive performance, interpretability, and
training time. To demonstrate the performance of these modeling
approaches and our pipeline on a case study, we used a data from a
previously published study that sought to classify individuals as having
healthy colons or colonic lesions based on the 16S rRNA gene sequences
collected from fecal samples (4). This dataset was selected because it
is a relatively large collection of individuals (N=490) connected to a
clinically significant disease where there is ample evidence that the
disease is driven by variation in the microbiome (2, 4, 5, 32). With
this dataset, we developed an ML pipeline that can be used in many
different scenarios for training and evaluating models. This framework
can be easily applied to other host-associated and environmental
microbiome datasets. We hope that this study \ldots{}

@@

We trained seven different models using the training data {[}Table 1{]}.
We focused on different classification algorithms and regularization
methods. Regularization helps to prevent overfitting by penalizing a
model that fits the training data too well (33). For regularized
logistic regression and SVM with a linear kernel, we used
L2-regularization to keep all potentially important features. For
comparison, we also trained an L1-regularized SVM with a linear kernel.
L1-regularization on microbiome data led to a sparser solution (i.e.,
forced many coefficients to zero). To explore the potential for
non-linear relationships among features to improve classification, we
trained tree-based models including, a decision tree, a random forest,
and gradient boosted trees (XGBoost) and an SVM with a non-linear
kernel.

@@

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performance of the seven models to
classify individuals as having healthy colons or SRNs {[}Figure 2{]}.
The predictive performance of random forest model was higher than other
ML models with a median 0.695 {[}IQR 0.650-0.739{]}, though not
significantly (p=0.5; The p-value was manually calculated using the
sampling distribution of the test statistic under the null hypothesis)
(Figure S3). Similarly, L2-regularized logistic regression, XGBoost,
L2-regularized SVM with linear and radial basis function kernel AUROC
values were not significantly different from one another and had median
AUROC values of 0.680 {[}IQR 0.639-0.750{]}, 0.679 {[}IQR
0.643-0.746{]}, 0.678 {[}IQR 0.639-0.750{]} and 0.668 {[}IQR
0.639-0.750{]}, respectively. L1-regularized SVM with linear kernel and
decision tree had significantly lower AUROC values than the other ML
models with median AUROC of 0.650 {[}IQR 0.629-0.760{]} and 0.601 {[}IQR
0.636-0.753{]}, respectively {[}Figure 2{]}. Interestingly, these
results demonstrate that the most complex model (XGBoost) did not have
the best performance and that the most interpretable models
(L2-regularized logistic regression and L2-regularized SVM with linear
kernel) performed nearly as well as non-linear models.

@@

\textbf{Interpretation of each ML model.} Interpretability is related to
the degree to which humans can understand the reasons behind a model
prediction (34--36). Because we often use ML models not just to predict
a health outcome, but also to identify potential biomarkers for disease,
model interpretation becomes crucial for microbiome studies. ML models
often decrease in interpretability as they increase in complexity. In
this study, we used two methods to help interpret our models.

@@

Second, to analyze non-linear models, we interpreted the feature
importance using permutation importance (37). Whereas the absolute
feature weights were determined from the trained models, here we
measured importance using the held-out test data. Permutation importance
analysis is a post hoc explanation of the model, in which we randomly
permuted groups of perfectly correlated features together and other
features individually across the two groups in the held-out test data.
We then calculated how much the predictive performance of the model
(i.e., testing AUROC values) decreased when each OTU or group of OTUs
was randomly permuted. We ranked the OTUs based on how much the median
testing AUROC decreased when it was permuted; the OTU with the largest
decrease ranked highest {[}Figure 4{]}. Among the twenty OTUs with the
largest impact, there was only one OTU (OTU 822) that was shared among
all of the models; however, we found three OTUs (OTUs 58, 110, 367) that
were important in each of the tree-based models. Similarly, the random
forest and XGBoost models shared four of the most important OTUs (OTUs
2, 12, 361, 477). Permutation analysis results also revealed that with
the exception of the decision tree model, removal of any individual OTU
had minimal impact on model performance. For example, if OTU 367 was
permuted across the samples in the decision tree model, the median AUROC
dropped from 0.601 to 0.525. In contrast, if the same OTU was permuted
in the random forest model, the AUROC only dropped from 0.695 to 0.680,
which indicated high degree of collinearity in the dataset. Permutation
analysis allowed us to gauge the importance of each OTU in non-linear
models and partially account for collinearity by grouping correlated
OTUs to determine their impact as a group.

@@

We highlighted the importance of model interpretation to gain greater
biological insights into microbiota-associated diseases. In this study,
we showcased two different interpretation methods: ranking each OTU by
(i) their absolute weights in the trained models and (ii) their impact
on the predictive performance based on permutation importance.
Human-associated microbial communities have complex correlation
structures that create collinearity in the datasets. This can hinder our
ability to reliably interpret models because the feature weights of
correlated OTUs are influenced by one another (38). To capture all
important features, once we identify highly ranked OTUs, we should
review their relationships with other OTUs. These relationships will
help us generate new hypotheses about the ecology of the disease and
test them with follow-up experiments. When we used permutation
importance, we partially accounted for collinearity by grouping
correlated OTUs to determine their impact as a group. We grouped OTUs
that had a perfect correlation with each other; however, we could reduce
the correlation threshold to further investigate the relationships among
correlated features. It is important to know the correlation structures
of the data to avoid misinterpreting the models. This is likely to be a
particular problem with shotgun metagenomic datasets where collinearity
will be more pronounced due to many genes being correlated with one
another because they come from the same chromosome. To identify the true
underlying microbial factors of a disease, it is crucial to follow up on
any correlation analyses with further hypothesis testing and
experimentation for biological validation.

In this study, we did not consider all possible modeling approaches.
However, the principles highlighted throughout this study apply to other
ML modeling tasks with microbiome data. For example, we did not evaluate
multicategory classification methods to predict non-binary outcomes. We
could have trained models to differentiate between people with healthy
colons and those with adenomas or carcinomas (k=3 categories). We did
not perform this analysis because the clinically relevant diagnosis
grouping was between patients with healthy colons and those with SRNs.
Furthermore, as the number of classes increases, more samples are
required for each category to train an accurate model. We also did not
use regression-based analyses to predict a non-categorical outcome. We
have previously used such an approach to train random forest models to
predict fecal short-chain fatty acid concentrations based on microbiome
data (39). Our analysis was also limited to shallow learning methods and
did not explore deep learning methods such as neural networks. Deep
learning methods hold promise (12, 40, 41), but microbiome datasets
often suffer from having many features and small sample sizes, which
result in overfitting.

Our framework provides a reproducible structure to investigators wanting
to train, evaluate, and interpret their own ML models to generate
hypotheses regarding which OTUs might be biologically relevant. However,
deploying microbiome-based models to make clinical diagnoses or
predictions is a significantly more challenging and distinct undertaking
(42). For example, we currently lack standardized methods to collect
patient samples, generate sequence data, and report clinical data. We
are also challenged by the practical constraints of OTU-based
approaches. The de novo algorithms commonly in use are slow, require
considerable memory, and result in different OTU assignments as new data
are added (43). Finally, we also need independent validation cohorts to
test the performance of a diagnostic model. To realize the potential for
using ML approaches with microbiome data, it is necessary that we direct
our efforts to overcome these challenges.

@@

\textbf{Data collection and study population.} The original stool
samples described in our analysis were obtained from patients recruited
by Great Lakes-New England Early Detection Research Network (5). Stool
samples were provided by adults who were undergoing a scheduled
screening or surveillance colonoscopy. Participants were recruited from
Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor
(MI, USA). Patients' colonic health was visually assessed by colonoscopy
with bowel preparation and tissue histopathology of all resected
lesions. We assigned patients into two classes: those with healthy
colons and those with screen relevant neoplasias (SRNs). The healthy
class included patients with healthy colons or non-advanced adenomas,
whereas the SRN class included patients with advanced adenomas or
carcinomas (44). Patients with an adenoma greater than 1 cm, more than
three adenomas of any size, or an adenoma with villous histology were
classified as having advanced adenomas (44). There were 172 patients
with normal colonoscopies, 198 with adenomas, and 120 with carcinomas.
Of the 198 adenomas, 109 were identified as advanced adenomas. Together
261 patients were classified as healthy and 229 patients were classified
as having an SRN.

\textbf{16S rRNA gene sequencing data.} Stool samples provided by the
patients were used for 16S rRNA gene sequencing to measure bacterial
population abundances. The sequence data used in our analyses were
originally generated by Baxter et al. (available through NCBI Sequence
Read Archive {[}SRP062005{]}, (5)). The OTU abundance table was
generated by Sze et al (45), who processed the 16S rRNA sequences in
mothur (v1.39.3) using the default quality filtering methods,
identifying and removing chimeric sequences using VSEARCH, and assigning
to OTUs at 97\% similarity using the OptiClust algorithm (43, 46, 47);
(\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
These OTU abundances were the features we used to predict colorectal
health of the patients. There were 6920 OTUs. OTU abundances were
subsampled to the size of the smallest sample and normalized across
samples such that the highest abundance of each OTU would be 1, and the
lowest would be 0.

@@

For hyperparameter selection, we started with a granular grid search.
Then we narrowed and fine-tuned the range of each hyperparameter.For
L2-regularized logistic regression, L1 and L2-regularized SVM with
linear and radial basis function kernels, we tuned the cost
hyperparameter, which controls the regularization strength, where
smaller values specify stronger regularization. For SVM with radial
basis function kernel, we also tuned the sigma hyperparameter, which
determines the reach of a single training instance where, for a high
value of sigma, the SVM decision boundary will be dependent on the
points that are closest to the decision boundary. For the decision tree
model, we tuned the depth of the tree where the deeper the tree, the
more splits it has. For random forest, we tuned the number of features
to consider when looking for the best tree split. For XGBoost, we tuned
the learning rate and the fraction of samples used for fitting the
individual base learners. Performing a grid search for hyperparameter
selection might not be feasible for when there are more than two
hyperparameters to tune for. In such cases, it is more efficient to use
random search or recently developed tools such as Hyperband to identify
good hyperparameter configurations (48).

@@

\hypertarget{ref-Quinn847475}{}
31. \textbf{Quinn TP}, \textbf{Erb I}. 2019. Another look at
microbemetabolite interactions: How scale invariant correlations can
outperform a neural network. bioRxiv.
doi:\href{https://doi.org/10.1101/847475}{10.1101/847475}.

\hypertarget{ref-knights_human-associated_2011}{}
32. \textbf{Knights D}, \textbf{Parfrey LW}, \textbf{Zaneveld J},
\textbf{Lozupone C}, \textbf{Knight R}. 2011. Human-associated microbial
signatures: Examining their predictive value. Cell Host Microbe
\textbf{10}:292--296.
doi:\href{https://doi.org/10.1016/j.chom.2011.09.003}{10.1016/j.chom.2011.09.003}.

\hypertarget{ref-Ng:2004:FSL:1015330.1015435}{}
33. \textbf{Ng AY}. 2004. Feature selection, l1 vs. l2 regularization,
and rotational invariance, pp. 78. \emph{In} Proceedings of the
twenty-first international conference on machine learning. ACM, New
York, NY, USA.

\hypertarget{ref-miller_explanation_2017}{}
34. \textbf{Miller T}. 2017. Explanation in artificial intelligence:
Insights from the social sciences. arXiv:170607269 {[}cs{]}.

\hypertarget{ref-ribeiro_why_2016}{}
35. \textbf{Ribeiro MT}, \textbf{Singh S}, \textbf{Guestrin C}. 2016.
``Why should i trust you?'': Explaining the predictions of any
classifier. arXiv:160204938 {[}cs, stat{]}.

\hypertarget{ref-nori_interpretml:_2019}{}
36. \textbf{Nori H}, \textbf{Jenkins S}, \textbf{Koch P},
\textbf{Caruana R}. 2019. InterpretML: A unified framework for machine
learning interpretability. arXiv:190909223 {[}cs, stat{]}.

\hypertarget{ref-10.1093ux2fbioinformaticsux2fbtq134}{}
37. \textbf{Altmann A}, \textbf{Toloşi L}, \textbf{Sander O},
\textbf{Lengauer T}. 2010. Permutation importance: a corrected feature
importance measure. Bioinformatics \textbf{26}:1340--1347.
doi:\href{https://doi.org/10.1093/bioinformatics/btq134}{10.1093/bioinformatics/btq134}.

\hypertarget{ref-dormann_collinearity:_2013}{}
38. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
\textbf{Buchmann C}, \textbf{Carl G}, \textbf{Carré G}, \textbf{Marquéz
JRG}, \textbf{Gruber B}, \textbf{Lafourcade B}, \textbf{Leitão PJ},
\textbf{Münkemüller T}, \textbf{McClean C}, \textbf{Osborne PE},
\textbf{Reineking B}, \textbf{Schröder B}, \textbf{Skidmore AK},
\textbf{Zurell D}, \textbf{Lautenbach S}. 2013. Collinearity: A review
of methods to deal with it and a simulation study evaluating their
performance. Ecography \textbf{36}:27--46.
doi:\href{https://doi.org/10.1111/j.1600-0587.2012.07348.x}{10.1111/j.1600-0587.2012.07348.x}.

\hypertarget{ref-sze_fecal_2019}{}
39. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
\textbf{Ruffin MT}, \textbf{Schloss PD}. 2019. Fecal short-chain fatty
acids are not predictive of colonic tumor status and cannot be predicted
based on bacterial community structure. mBio \textbf{10}:e01454--19.
doi:\href{https://doi.org/10.1128/mBio.01454-19}{10.1128/mBio.01454-19}.
